On the reality of the quantum state once again: A no-go theorem for $\psi$-ontic models

### 2023-02-16 Articles/2021-HarriganSpekkens/HarriganSpekkensTakeTwo.tex
HarriganSpekkens: minor revisions from Andrea

\begin{abstract}
	In this paper we show that $\psi$-ontic models, as defined by Harrigan and Spekkens (HS), cannot reproduce quantum theory. Instead of focusing on probability, we use information theoretic considerations to show that all pure states of $\psi$-ontic models must be orthogonal to each other, in clear violation of quantum mechanics. Given that  (1) Pusey, Barrett and Rudolph (PBR) previously showed that $\psi$-epistemic models, as defined by HS, also contradict quantum mechanics, and (2) the HS categorization is exhausted by these two types of models, we conclude that the HS categorization itself is problematic as it leaves no space for models that can reproduce quantum theory.
\end{abstract}

@@

While the discussion has been focused mainly on whether $\psi$-epistemic models are problematic or whether the PBR argument itself is problematic, we ask a different question: what if the underlying HS classification itself is problematic? In particular, the framework assumes that once pure states are fully characterized, mixed states are constructed simply as a standard (i.e. classical) statistical mixture. But classical mixtures and quantum mixtures have different properties. For example, classical mixtures have a single decomposition in terms of pure states while quantum mixtures do not. Studying the behavior of states under mixing, then, provides a new way to explore possible violations.

Using this new approach, in this paper we are going to provide a no-go theorem for $\psi$-ontic models. The core insight is that the degree of overlap between two probability distributions affects how the entropy of their mixture relates to the entropy of the components. Consequently, we find that the lack of overlap required by $\psi$-ontic models forces all pure states to be orthogonal. Since this directly contradicts quantum theory, we conclude that $\psi$-ontic models are not viable.

@@

The exposition of the HS classification makes it seem one has complete freedom in choosing the overlap between epistemic states. In this section we will show that this is not the case: this is constrained by entropy of mixtures. These constraints cannot be replicated by a $\psi$-ontic model with standard information theoretic techniques. Since quantum statistical mechanics and quantum information theory depend on the correct value of entropy on mixtures, $\psi$-ontic models will violate quantum theory.

To show the violation, we will first study how the entropy behaves when mixing two states, both for the Shannon/Gibbs entropy and the von Neumann entropy.\footnote{Full calculations with all steps are included in the appendix.} The first will be used to calculate entropy of statistical mixtures of epistemic states, since equation \ref{epistemic_mixing} tells us the mixture is done according to a standard (classical) measure of probability. The second will be used to calculate the entropy on the quantum case.

@@

\emph{Proof}: Suppose we have a $\psi$-ontic model defined according to \cite{Harrigan:2010}. The epistemic states $p(\lambda|P_\psi)$ and $p(\lambda|P_\phi)$ consist of non-overlapping probability distributions over a space $\Lambda$. That is, they are two non-overlapping probability density functions defined to be the Radon-Nikodym derivatives between the respective probability measures over $\Lambda$ and the measure $\lambda$. Equation \ref{epistemic_mixing} tells us that the probability measures combine in the usual way, and therefore an equal mixture of the two states must obey equation \ref{entropy_nonoverlap}. Given that $\psi$ and $\phi$ are pure states, and the entropy for pure states in quantum mechanics is zero, we must have
\begin{equation}\label{entropy_pure}
	H(p(\lambda|P_\psi)) = H(p(\lambda|P_\phi)) = 0,
\end{equation}
and therefore
\begin{equation}\label{required_entropy}
	\begin{aligned}
		H\left(p(\lambda|\frac{1}{2}P_\psi + \frac{1}{2}P_\phi)\right) &= \\
		H\left(\frac{1}{2}p(\lambda|P_\psi) + \frac{1}{2}p(\lambda|P_\phi)\right) 
		&= 1.
	\end{aligned}
\end{equation}
If we compare the above with eq.~\ref{entropy_mixed}, it follows that $p$ must be zero. In fact, the only case in which the Shannon entropy for a two-element distribution is equal to one is for the uniform distribution $H(1/2, 1/2) = 1$.  That is, we must have that
\begin{equation}\label{orthogonal}
	\langle \psi | \phi \rangle = 0
\end{equation}
no matter what $\psi$ and $\phi$ are.

@@

First of all, information entropy is strictly concave, meaning that $H(p \rho_1 + (1-p) \rho_2) \geq p H(\rho_1) + (1-p) H(\rho_2)$ with the equality valid only if $\rho_1$ and $\rho_2$ are the same exact state. That is, we can check whether two states $\rho_1$ and $\rho_2$ are the same state simply by creating a mixture $\rho = \frac{1}{2} \rho_1 + \frac{1}{2} \rho_2$ and see whether we end up with the same entropy or not. In other words: \emph{entropy fixes equality}. %spiega meglio
The issue here is that the mapping between epistemic states and density matrices is not injective. Quantum mixtures, in fact, do not have a single decomposition in pure states. For example, the maximally mixed state of a qubit for a spin-$1/2$ system can be achieved with either an equal mixture of spin up and down or an equal mixture of spin left and right. Therefore epistemic states cannot be mapped one-to-one to quantum density operators, but rather each density operator corresponds to an equivalence class of epistemic states.\footnote{The degree of degeneracy one has to require is also extreme. For a qubit, in the simplest case, we would have a $\psi$-ontic state for each direction on the Bloch sphere, meaning the space of ontic states is the sphere $S^2$. The space of all possible epistemic states is the set of all possible distributions integrable to one. This is an infinite dimensional function space. However, the space of mixed states for a quibit is the interior of the Bloch sphere, which is a bounded three-dimensional manifold. The equivalence classes of epistemic states, then, must be infinite dimensional.} This degeneracy means that the notion of equality among epistemic states is different than the notion of equality among density operators. In turn, this means that whatever definition we have for entropy on epistemic states must not satisfy strict concavity, if we want the value of entropy to agree between epistemic states and density operators. \emph{Entropy cannot fix equality for epistemic states}. In other words, one can make mixtures of different epistemic states without increasing entropy, which is questionable on physics grounds.

@@

If we combine our result with PBR, the only models that are not ruled out are $\psi$-epistemic models where the objects are not independently prepared. That is, where the joint distribution over ontic states is not the product of the marginals. However, this loophole can also be closed with information theoretic considerations. It is a well-known result in both classical and quantum information theory that the entropy of a joint distribution is the sum of the marginal if and only if the subsystems are independent (\cite{Ash:2010}, \cite{Nielsen:2010}). Therefore an epistemic state that represents independent quantum states must also be the joint distribution of statistically independent epistemic states of the individual systems. This tells us that $\psi$-epistemic models must represent composite states of independently prepared systems with independently distributed epistemic states, and therefore PBR applies. This closes the last loophole: no model in the HS categorization can reproduce quantum theory in its entirety, which includes quantum information and quantum statistical mechanics.

Thus, one concludes the following:
\begin{quote}
	\textbf{No ontological model can reproduce quantum theory}.
\end{quote}
Now, the result may seem, at first, preposterous. But once we analyze the mathematics in detail, we see that both results, ours and PBR, are fundamentally finding something, in retrospect, obvious.

@@

By using standard information theoretic techniques, we have shown that $\psi$-ontic models are not compatible with quantum information theory, and therefore with quantum mechanics itself. Given that the PBR theorem already ruled out $\psi$-epistemic models, the HS categorization of ontological models is fundamentally empty. Thus, such a framework should be employed neither to classify quantum interpretations, nor to draw conclusions on the nature of the quantum state.

The key problem of the HS categorization is that it is ``not quantum enough.'' The framework tries to hide the non-classicality at the level of each ontic state, which, by construction, cannot be decomposed further using a standard probability measure. However, it posits that statistical mixtures are simply classical mixtures in the sense that they are captured by a standard probability measure. But quantum mixtures are non-classical in the same way that the quantum transition probabilities defined by the inner product are non-classical. Therefore the HS categorization prevents one from describing quantum statistical mixing.

Our conclusion is that the PBR theorem finds that $\psi$-epistemic models are not sustainable not because there is a problem with $\psi$-epistemic models in particular, but rather there is an underlying problem in the HS definitions. Specifically, the use of standard probability theory in composite systems. If understood in this manner, the PBR result is the first part of a more general argument showing the inadequacy of the HS framework to categorize quantum theories, an argument that we have completed with the proposed no-go theorem against $\psi$-ontic models.

>>>>

\begin{abstract}
	In this paper we show that $\psi$-ontic models, as defined by Harrigan and Spekkens (HS), cannot reproduce quantum theory. Instead of focusing on probability, we use information theoretic considerations to show that all pure states of $\psi$-ontic models must be orthogonal to each other, in clear violation of quantum mechanics. Given that (i) Pusey, Barrett and Rudolph (PBR) previously showed that $\psi$-epistemic models, as defined by HS, also contradict quantum mechanics, and (ii) the HS categorization is exhausted by these two types of models, we conclude that the HS categorization itself is problematic as it leaves no space for models that can reproduce quantum theory.
\end{abstract}

@@

While the discussion has been focused mainly on whether $\psi$-epistemic models are problematic or whether the PBR argument itself is problematic, we ask a different question: what if the underlying HS classification itself is problematic? In particular, the framework assumes that once pure states are fully characterized, mixed states are constructed simply as a standard (i.e. classical) statistical mixture. However, as is well-known, classical and quantum mixtures have different properties. For instance, classical mixtures have a single decomposition in terms of pure states while quantum mixtures do not. Studying the behavior of states under mixing, then, provides a new way to explore \textcolor{blue}{possible violations}. %of what? I would also mention that the example will be relevant for our argument.

\textcolor{blue}{Using this new approach}, %which new approach? Io utilizzerei la frase che era presente nella versione precedente: ``using information theoretic considerations we show that $\psi$-ontic models cannot reproduce all the predictions, results and implications of QM since the lack of overlap of epistemic states---a necessary condition for a model to be $\psi$-ontic---requires that all states must be orthogonal, which openly contradicts quantum theory.''
in this paper we are going to provide a no-go theorem for $\psi$-ontic models. \textcolor{red}{The core insight is that the degree of overlap between two probability distributions affects how the entropy of their mixture relates to the entropy of the components.} %contrary to some previous version of the paper, this sentence is not put in context and some readers not familiar with HS will not understand. 
Consequently, we find that the lack of overlap required by $\psi$-ontic models forces all pure states to be orthogonal. Since this directly contradicts quantum theory, we conclude that $\psi$-ontic models are not viable.

@@

%The exposition of the: ci possono obiettare che  la nostra esposizione a non essere fedele, quando invece  proprio il framework di HS d avere il problema di cui parliamo.
Looking at the HS classification it seems that one has complete freedom in choosing the overlap between epistemic states. %for argumentative strategies I would explain why it seems so.
In this section we will show that this is not the case: this choice is in fact constrained by entropy of mixtures. Moreover, these constraints cannot be replicated by a $\psi$-ontic model with standard (i.e.\ classical) information theoretic techniques. Since quantum statistical mechanics and quantum information theory depend on the correct value of entropy on mixtures, $\psi$-ontic models will violate quantum theory. %to make our argument stronger we should show that psi-ontic models rely on classical information theoretic structures. This fact should clearly stated now not in the next paragraph.

To show the violation, we will first study how the entropy behaves when mixing two states, both for the Shannon/Gibbs entropy and the von Neumann entropy.\footnote{Full calculations with all steps are included in the appendix.} The first will be used to calculate entropy of statistical mixtures of epistemic states, since equation \ref{epistemic_mixing} tells us the mixture is done according to a standard (classical) measure of probability. The second will be used to calculate the entropy on the quantum case. % always for the sake of clarity, here I would anticipate what we are going to say: given the formal structure on which psi-ontic models are built they cannot reproduce QM. In this way the argument will be already "in the head" of our readers.

@@

\emph{Proof}: Suppose we have a $\psi$-ontic model defined according to \cite{Harrigan:2010}. The epistemic states $p(\lambda|P_\psi)$ and $p(\lambda|P_\phi)$ consist of non-overlapping probability distributions over a space $\Lambda$. That is, they are two non-overlapping probability density functions defined to be the Radon-Nikodym derivatives between the respective probability measures over $\Lambda$ and the measure $\lambda$. Equation \ref{epistemic_mixing} tells us that the probability measures combine in a classical way, and therefore an equal mixture of the two states must obey equation \ref{entropy_nonoverlap}. Given that $\psi$ and $\phi$ are pure states, and the entropy for pure states in quantum mechanics is zero, we must have
\begin{equation}\label{entropy_pure}
	H(p(\lambda|P_\psi)) = H(p(\lambda|P_\phi)) = 0,
\end{equation}
and therefore
\begin{equation}\label{required_entropy}
	\begin{aligned}
		H\left(p(\lambda|\frac{1}{2}P_\psi + \frac{1}{2}P_\phi)\right) &= \\
		H\left(\frac{1}{2}p(\lambda|P_\psi) + \frac{1}{2}p(\lambda|P_\phi)\right) 
		&= 1.
	\end{aligned}
\end{equation}
If we compare the above with eq.~\ref{entropy_mixed}, it follows that $p$ must be zero. In fact, the only case in which the Shannon entropy for a two-element distribution is equal to one is for the uniform distribution $H(1/2, 1/2) = 1$.  That is, we must have that
\begin{equation}\label{orthogonal}
	\langle \psi | \phi \rangle = 0
\end{equation}
no matter what $\psi$ and $\phi$ are.

@@

First of all, information entropy is strictly concave, meaning that $H(p \rho_1 + (1-p) \rho_2) \geq p H(\rho_1) + (1-p) H(\rho_2)$ with the equality valid only if $\rho_1$ and $\rho_2$ are the same exact state. That is, we can check whether two states $\rho_1$ and $\rho_2$ are the same state simply by creating a mixture $\rho = \frac{1}{2} \rho_1 + \frac{1}{2} \rho_2$ and see whether we end up with the same entropy or not. In other words: \emph{entropy fixes equality}.
The issue here is that the mapping between epistemic states and density matrices is not injective. Quantum mixtures, in fact, do not have a single decomposition in pure states. For example, the maximally mixed state of a qubit for a spin-$1/2$ system can be achieved with either an equal mixture of spin up and down or an equal mixture of spin left and right. Therefore epistemic states cannot be mapped one-to-one to quantum density operators, but rather each density operator corresponds to an equivalence class of epistemic states.\footnote{The degree of degeneracy one has to require is also extreme. For a qubit, in the simplest case, we would have a $\psi$-ontic state for each direction on the Bloch sphere, meaning the space of ontic states is the sphere $S^2$. The space of all possible epistemic states is the set of all possible distributions integrable to one. This is an infinite dimensional function space. However, the space of mixed states for a quibit is the interior of the Bloch sphere, which is a bounded three-dimensional manifold. The equivalence classes of epistemic states, then, must be infinite dimensional.} This degeneracy means that the notion of equality among epistemic states is different than the notion of equality among density operators. In turn, this means that whatever definition we have for entropy on epistemic states must not satisfy strict concavity, if we want the value of entropy to agree between epistemic states and density operators. \emph{Entropy cannot fix equality for epistemic states}. In other words, one can make mixtures of different epistemic states without increasing entropy, which is questionable on physics grounds.

@@

If we combine our result with the PBR theorem, the only models that are not ruled out by the HS framework are $\psi$-epistemic models in which quantum objects are not independently prepared. That is, where the joint distribution over ontic states is not the product of the marginals. However, this loophole can also be closed with information theoretic considerations. It is a well-known result in both classical and quantum information theory that the entropy of a joint distribution is the sum of the marginal if and only if the subsystems under consideration are independent (\cite{Ash:2010}, \cite{Nielsen:2010}). Therefore an epistemic state that represents independent quantum states must also be the joint distribution of statistically independent epistemic states of the individual systems. This tells us that $\psi$-epistemic models must represent composite states of independently prepared systems with independently distributed epistemic states, and therefore PBR applies. This closes the last loophole: no model in the HS categorization can reproduce quantum theory in its entirety, which includes quantum information and quantum statistical mechanics.

Thus, one concludes the following:
\begin{quote}
	\textbf{No ontological model can reproduce quantum theory}.
\end{quote}
Although this result may at first sight seem paradoxical, once we analyze in detail the mathematics behind HS categorization we see that both results---ours and PBR---are finding something, in retrospect, self-evident.

@@

By using standard information theoretic techniques, we have shown that $\psi$-ontic models are not compatible with quantum information theory, and therefore with quantum mechanics itself. Given that the PBR theorem already ruled out $\psi$-epistemic models, the HS categorization of ontological models is found to be fundamentally empty. Thus, such a framework should be employed neither to classify quantum interpretations, nor to draw conclusions on the nature of the quantum state.

The key problem of the HS categorization is that it is ``not quantum enough.'' The framework tries to hide the non-classicality at the level of each ontic state, which, by construction, cannot be decomposed further using a standard probability measure. However, HS definitions entail that statistical mixtures are simply classical mixtures in the sense that they are captured by a standard probability measure. But quantum mixtures are non-classical in the same way that the quantum transition probabilities defined by the inner product are non-classical. Therefore, the HS categorization prevents one from describing quantum statistical mixing.

Our conclusion is that the PBR theorem finds that $\psi$-epistemic models are not sustainable because of an underlying issue of the HS definitions, not because epistemic interpretations contradict quantum theory. Specifically, the problem of HS approach is the use of standard probability theory in composite systems. If understood in this manner, the PBR result is the first part of a more general argument showing the inadequacy of the HS framework to categorize quantum theories, an argument that we have completed with the proposed no-go theorem against $\psi$-ontic models.
